Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-02-22 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Add analysis model impact data

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: Johanna Mu\~noz Avila
    affiliation: University Medical Center Utrecht \AND
  - name: Valentijn de Jong
    affiliation: University Medical Center Utrecht 
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This is a tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Footnotes in the current version show work in progress/under construction. The last section is not part of the manuscript, but purely for reminders.
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article: default
    word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

# packages
library(tidyverse)
library(mice)
library(miceadds)
# library(metamisc)
# library(GJRM)
# library(pan)

@@

  - `name` Name of the study,
  - `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
  - `age` Age of the patient,
  - `motor_score` Glasgow Coma Scale motor score,
  - `pupil` Pupillary reactivity,
  - `ct` Marshall Computerized Tomography classification,
  - `hypox` Hypoxia (0=no, 1=yes),
  - `hypots` Hypotension (0=no, 1=yes),
  - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
  - `edh` Epidural hematoma (0=no, 1=yes),
  - `mort` 6-month mortality (0=alive, 1=dead).

The analysis model for this dataset is a prediction model with `mort` as the outcome.^[Look at analysis model, maybe copy from GREAT data example e.g., adjusted prognostic effect of ct on unfortunate outcomes. Add something about systematically missing data here.] The data is already imputed (Steyerberg et al, 2008), so we've induced missingness again based on the missingness in the original data.

```{r impact, echo=FALSE, fig.width=6, fig.height=8}
# load data
data("impact", package = "metamisc")
impact_NA <- readRDS("../Data/impact_NA.RDS")
# # descriptive statistics
# by(impact, impact$name, summary) 
# psych::describe(impact)[,c(2:5,8:9)]
# # missingness
plot_md_pat(impact_NA)
# TODO: rotate variable labels
# TODO: think about nr of clusters: if there's too many, faceting is no option -> some distribution across the clusters like shepley plot
# TODO: make missingness % gradient for clusters (ie the number of clusters for which this variable is missing).

@@

```{r echo=FALSE}
# md pattern in first cluster
plot_md_pat(impact_NA[impact_NA$name == "TINT", ])
plot_md_pat(impact_NA[impact_NA$name == "APOE", ])
```

>>>>

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: Johanna Mu\~noz Avila
    affiliation: University Medical Center Utrecht \AND
  - name: Valentijn de Jong
    affiliation: University Medical Center Utrecht 
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This is a tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Footnotes in the current version show work in progress/under construction. The last section is not part of the manuscript, but purely for reminders. We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables").
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article: default
    word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

# packages
library(tidyverse)
library(mice)
library(miceadds)
library(lme4)
library(broom.mixed)
# library(metamisc)
# library(GJRM)
# library(pan)

@@

  - `name` Name of the study,
  - `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
  - `age` Age of the patient,
  - `motor_score` Glasgow Coma Scale motor score,
  - `pupil` Pupillary reactivity,
  - `ct` Marshall Computerized Tomography classification,
  <!-- - `hypox` Hypoxia (0=no, 1=yes), -->
  <!-- - `hypots` Hypotension (0=no, 1=yes), -->
  <!-- - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes), -->
  <!-- - `edh` Epidural hematoma (0=no, 1=yes), -->
  - `mort` 6-month mortality (0=alive, 1=dead).

The analysis model for this dataset is a prediction model with `mort` as the outcome.^[Look at analysis model, maybe copy from GREAT data example e.g., adjusted prognostic effect of `ct` on unfortunate outcomes, we just want to know the adjusted odds ratio for `ct`. Add something about systematically missing data here.] The data is already imputed (Steyerberg et al, 2008), so we've induced missingness again based on the missingness in the original data.

$$
\text{mort}_{ij} =
\gamma_{00} + 
\gamma_{01} \text{ type}_{j} + 
\gamma_{10} \text{ age}_{ij} + 
\gamma_{10} \text{ moter\_score}_{ij} + 
\gamma_{10} \text{ pupil}_{ij} + 
\gamma_{10} \text{ ct}_{ij} + \\
u_{0j} + 
u_{1j} \text{ age}_{ij} + 
u_{1j} \text{ moter\_score}_{ij} +  
u_{1j} \text{ pupil}_{ij} +  
u_{1j} \text{ ct}_{ij} +  
e_{ij} \\
$$

`glmer(mort ~ 1 + type + age + motor_score + pupil + ct + hypox + hypots + tsah + edh + (1 | name))`

<!-- \gamma_{10} \text{ hypox}_{ij} + -->
<!-- \gamma_{10} \text{ hypots}_{ij} + -->
<!-- \gamma_{10} \text{ tsah}_{ij} + -->
<!-- \gamma_{10} \text{ edh}_{ij} + -->

<!-- u_{1j} \text{ hypox}_{ij} + -->
<!-- u_{1j} \text{ hypots}_{ij} + -->
<!-- u_{1j} \text{ tsah}_{ij} + -->
<!-- u_{1j} \text{ edh}_{ij} + -->

```{r impact, echo=FALSE, fig.width=6, fig.height=8}
# load data
data("impact", package = "metamisc")
impact_NA <- readRDS("../Data/impact_NA.RDS") 
# # descriptive statistics
# by(impact, impact$name, summary) 
# psych::describe(impact)[,c(2:5,8:9)]
# # missingness
plot_md_pat(impact_NA)
# TODO: rotate variable labels
# TODO: think about nr of clusters: if there's too many, faceting is no option -> some distribution across the clusters like shepley plot
# TODO: make missingness % gradient for clusters (ie the number of clusters for which this variable is missing).
```

```{r impact_md, echo=FALSE}
# md pattern in first cluster
plot_md_pat(impact_NA[impact_NA$name == "TINT", ])
plot_md_pat(impact_NA[impact_NA$name == "APOE", ])
```

```{r impact_cca}
mod_cca <- glmer(mort ~ type + age + motor_score + pupil + ct + (1 | name), family = "binomial", data = impact_NA) 
mod_cca %>% broom.mixed::tidy()

@@

```{r impact_pred}
quickpred(impact_NA)
```
